---
title: "Modeling Workspace"
author: "Will Morgan"
date: "October 3, 2018"
output: pdf_document
---

- pick a goddamn sample already
- estimate poisson models
- check for possible over- or under-dispersion
- estimate QP models
- compare aic of nb and pois models (to find which formula specification is best, and to show if nb > pois)
- compare estimates and std errors of qp, nb models
- show model fit statistics (std res, influence plot, error function)
MAYBE <!-- - run zip/zinb -->
MAYBE <!-- - compare fit of zip/zinb with AIC/err func (show model fit stats) -->
MAYBE <!-- - compare estimates and std errors of nb, zip/zinb model -->                    
- calc/interpret av mfx


### 0. Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

## 0. Setup and Import
libs <- c("data.table", "dplyr", "magrittr", "ggplot2", "forcats", "purrr", "AER",
          "pscl", "broom", "here", "margins")
lapply(libs, library, character.only = TRUE)
source(here("Code/utils/model_utils.R"))

data <- fread(here("Data/thread_cleaned.csv")) %>%
  mutate(faculty_owner = if_else(thread_owner_role == "P", "Y", "N"),
         faculty_responded = if_else(faculty_responses > 0, "Y", "N")) %>%
  as.data.table() 
```

### 1. Sample Selection
```{r sample}
## 1. Sample selection (filters)

# 99% quantile filter
data %<>%
  filter(thread_participants <= 9,
         thread_length <= 16,
         student_responses <= 15,
         studs_in_thread <= 9,
         parent_views <= 54,
         !is.na(parent_views),
         !is.na(thread_width)) %>%
  filter(strm %in% c(2167, 2171, 2177, 2181)) %>%
  as.data.table()

# # 99.5% quantile filter
# data %<>%
#   filter(thread_participants <= 15,
#          thread_length <= 32,
#          student_responses <= 30,
#          studs_in_thread <= 15,
#          !is.na(parent_views),
#          !is.na(thread_width),
#          parent_views <= 125) %>%
#   filter(strm %in% c(2167, 2171, 2177, 2181)) %>%
#   as.data.table()


#### Potential subsets

## Course sections with >500 threads
# sections <- data %>%
#   group_by(bb_course_id) %>%
#   summarise(posts = sum(thread_length),
#             threads = n()) %>%
#   filter(threads >= 500) %>%
#   select(bb_course_id) %>%
#   unlist()
# 
# nsr <- data[bb_course_id %in% sections, .(student_responses, faculty_responded, faculty_owner,
#                 parent_views, parent_post_length)][
#                   , parent_post_length := log(1 + parent_post_length)]


## Subjects known to have high forum use
# subjects <- c("ENG", "PSY", "COM", "BIS", "OGL")
# subjects <- c("ENG")
# nsr <- data[subject %in% subjects, .(student_responses, faculty_responded, faculty_owner,
#                 parent_views, parent_post_length)][
#                   , parent_post_length := log(1 + parent_post_length)]

###NOTES:
# ENG, PSY, COM, BIS, OGL are top 5 (in order) most occurring subjects

## Courses known to have high forum use
courses <- data %>%
  group_by(course) %>%
  summarise(posts = sum(thread_length),
            threads = n()) %>%
  ungroup() %>%
  arrange(desc(posts))

head(courses, 10)

courses %<>%
  slice(1:5) %>%
  select(course) %>%
  unlist()

nsr <- data[course %in% courses, .(student_responses, faculty_responded, faculty_owner,
                                  parent_views, parent_post_length)][
                                    , parent_post_length := log(1 + parent_post_length)]

# nsr <- data[course == "ENG102", .(student_responses, faculty_responded, faculty_owner,
#                                   parent_views, parent_post_length)][
#                                     , parent_post_length := log(1 + parent_post_length)]

```

### 2. Model Setup
```{r formulas}
# setup
formulas <- list(
  as.formula(student_responses ~ faculty_responded),
  as.formula(student_responses ~ faculty_owner),
  as.formula(student_responses ~ .),
  as.formula(student_responses ~ . + faculty_owner*faculty_responded +
               faculty_owner*parent_views + faculty_owner*parent_post_length),
  as.formula(student_responses ~ (.)^2)
)

formula_names <- c("Null", "Faculty Responses Only", "Faculty Owner Only",
                   "All Predictors", "Predictors and Interactions", "Saturated")

#nsr %<>% sample_frac(0.1)

```


### 3. Poisson Models and testing for equidispersion
```{r poisson, echo = TRUE}
pois <- runCountMods(formulas, nsr, "poisson")
names(pois) <- formula_names

# Dispersion (ratio of deviance to residual.df, deviance approaches chi2 with mean = df)
map(pois, function(x) x$deviance / x$df.residual)
map(pois, dispersiontest, alternative = 'two.sided')
```


### 4. QP Models
```{r qp, echo = TRUE}
## QP
qp <- runCountMods(formulas, nsr, 'quasipoisson')
names(qp) <- formula_names

summary(qp[[3]])
summary(qp[[4]])

```


qp dispersion estimates are strangely above and below 1 - inconsistent measure
of dispersion, but we can see that the models with more vars are consistenly
underdispersed. Since we can assume that these models are better fitting (AIC, LR test)
then the dispersion estimate is probably better. Move forward assuming that 
we have underdispersed data (which rules out use of NB for now)


### 5a. NB models (run if overdispersed)
```{r nb, echo = TRUE}
nb <- runCountMods(formulas, nsr, family = 'nb')
names(nb) <- formula_names

summary(nb[[3]])
summary(nb[[4]])

# nb_test1 <- MASS::glm.nb(formulas[[1]], nsr,
#                             control = glm.control(maxit = 250, trace = 3))
# 
# nb_test2 <- MASS::glm.nb(formulas[[3]], nsr,
#                             control = glm.control(maxit = 250, trace = 3))

```


if we're using zip we need to come up with some sort of reason as to why there might
be a structural component of the distribution creating zeros:

question:

is there an underlying process that is generating a higher-than-expected amount
of threads with no student responses?

possible answer:

participants are creating threads not intended for actual discussion. Instructors
might do this to disseminate a bunch of info at once (syllabus, project descr., etc.)
or another scenario might be that they create a "stickied" thread describing 
what the students need to post in that forum. Might be happening with students
as well if instructors are using forums to just collect responses to a discussion 
prompt

next step:

if we're going to assume that the above answer is the reason for why we are seeing
so many zeros, how can we identify this? what would be a good predictor for this happening?

- faculty responded (if student posts a question answered by a faculty, no need for student to reply)
- faculty owner (for reasons described above)
- parent views (threads like the one described above would likely be meant for a wide audience, hence more views)
- parent post length (meh)
- faculty_owner*parent_views (if made by student and has low views, probably less chance of being responded to)


### 5b. ZIP Models (run if underdispersed - structural zeros)
```{r zip, echo = TRUE}
library(pscl)

zip_test1 <- zeroinfl(formulas[[1]], nsr, dist = 'poisson', link = 'logit')
summary(zip_test)

zip_test2 <- zeroinfl(formulas[[3]], nsr, dist = 'poisson', link = 'logit')
summary(zip_test2)

# full desired model
formula <- as.formula("student_responses ~ faculty_responded + faculty_owner +
                      parent_views + parent_post_length + faculty_owner*faculty_responded +
                      faculty_owner*parent_views + faculty_owner*parent_post_length | faculty_responded + faculty_owner + parent_views + parent_post_length + faculty_owner*parent_post_length  + faculty_owner*parent_views")

zip_test2 <- zeroinfl(formula, nsr, dist = 'poisson', link = 'logit')
summary(zip_test2)

```



### 6. Model Comparisons

#### AIC Check
```{r}
## Pois and NB AICs
aic <- bind_rows(
  data.frame(aic = map_dbl(nb, AIC), model = "nb", formula = names(nb), stringsAsFactors = FALSE),
  data.frame(aic = map_dbl(pois, AIC), model = "pois", formula = names(pois), stringsAsFactors = FALSE)
)

ggplot(aic, aes(factor(formula, levels = rev(formula_names)), aic - min(aic), fill = model)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  coord_flip() +
  theme(axis.title.y = element_blank())

# lrtest of NB saturated and NB predictors and interactions ====> use saturated NB as final model
lrtest(nb[[6]], nb[[5]])

```

#### Estimate Comparisons

```{r no interactions}
# all vars, no interactions
knitr::kable(tidy(qp[[4]]))
knitr::kable(tidy(nb[[4]]))

# # all vars, some interactions
# knitr::kable(tidy(qp[[5]]))
# knitr::kable(tidy(nb[[5]]))
# 
# # saturated
# knitr::kable(tidy(qp[[6]]))
# knitr::kable(tidy(nb[[6]]))

## all var no interaction estimate plot
qp_4 <- tidy(qp[[4]]) %>%
  mutate(method = "QP")

nb_4 <- tidy(nb[[4]]) %>%
  mutate(method = "NB")

bind_rows(qp_4, nb_4) %>%
  ggplot(aes(term, estimate, fill = factor(method))) +
  geom_bar(position = 'dodge', stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  coord_flip()

```

```{r some interactions} 
## all var no interaction estimate plot
qp_5 <- tidy(qp[[5]]) %>%
  mutate(method = "QP")

nb_5 <- tidy(nb[[5]]) %>%
  mutate(method = "NB")

bind_rows(qp_5, nb_5) %>%
  ggplot(aes(term, estimate, fill = factor(method))) +
  geom_bar(position = 'dodge', stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  coord_flip()

```


#### 7. Model diagnostics

```{r influence plots, eval = F}
# Extract OG model
nsr0 <- nsr
model0 <- nb[[4]]
influencePlot(model0)

sr0 <- MASS::studres(model0)
sum(abs(sr0) > 2)                  # number of obs with high studentized residuals
filter <- abs(sr0) <= 2

# Create new data and run again
nsr1 <- nsr[filter, ]
model1 <- MASS::glm.nb(formulas[[3]], data = nsr1)
influencePlot(model1)

sr1 <- MASS::studres(model1)
sum(abs(sr1) > 2)                  # number of obs with high studentized residuals
filter <- abs(sr1) <= 2

# Another one
nsr2 <- nsr1[filter, ]
model2 <- MASS::glm.nb(formulas[[3]], data = nsr2)
influencePlot(model2)
```


### 8. Average Marginal Effects
```{r calculate MEs}

## Cannot be data.table for margins to work
nsr %<>% as.data.frame()

model <- MASS::glm.nb(formulas[[5]], data = nsr)

# pure margins
mfx <- margins(model, type = 'response')
summary(mfx)

# at different number of views
mfx_views <- summary(margins(model, type = 'response', at = list(parent_views = seq(0, 50, 10))))

# at different lengths
mfx_lengths <- summary(margins(model, type = 'response', at = list(parent_post_length = seq(0, 10, 1))))
```

```{r plot main effects}
main_ame <- summary(mfx)

ggplot(main_ame, aes(factor, AME)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.3) +
  ylim(c(-1, 1)) +
  geom_hline(yintercept = 0, linetype = 'dashed') +
  labs(title = "Average Effect of a 1-unit increase",
       y = "Change in # of Student Responses",
       x = "Variable") +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r plot fx at levels}
ggplot(mfx_views, aes(parent_views, AME, color = factor(factor))) + 
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin = lower, ymax = upper, width = 0.25)) +
  labs(title = "Effect of a 1-unit increase in Theoretical Posts with X Views",
       x = "Number of Views",
       y = "Change in number of student responses") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.title = element_blank())

ggplot(mfx_lengths, aes(parent_post_length, AME, color = factor(factor))) + 
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin = lower, ymax = upper, width = 0.25)) +
  labs(title = "Effect of a 1-unit increase in Theoretical Posts with X Views",
       x = "ln(Post Length)",
       y = "Change in number of student responses") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.title = element_blank())


```















